---
title: "Pathways of potential responder genes"
author: "Christina Hillig"
date: "7/25/2022"
output: 
  pdf_document: default
  html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = TRUE)
```

# Introduction
The following section describes the pathway enrichment analysis of up-regulated genes in the IL17A transcript positive and negative group in Leukocytes in the ST data set. 

# Initialisation
Load required R and Bioconductor packages:
```{r, message=FALSE}
# R packages
rlibs <- c("dplyr", "gtools", "hash", "kableExtra", "knitr", 'GOplot',
           "stringr", "tibble", "xlsx", "hash", "Hmisc", 'qdapTools', 'msigdbr')
invisible(lapply(rlibs, require, character.only = TRUE))
# Bioconductor packages
bioclibs <- c("ReactomePA", "pathview",  "enrichplot", "org.Hs.eg.db", "DOSE")
invisible(lapply(bioclibs, require, character.only = TRUE))

getwd()
environment()

# Set seed for reproduceability 
set.seed(42)

# Source R-scripts
source(file.path("..", "..", "..", "r_scripts", "pathway_analysis", 'init.R'))
source(file.path("..", "..", "..", "r_scripts", "pathway_analysis", 'load_data.R'))
source(file.path("..", "..", "..", "r_scripts", "pathway_analysis",'utils.R'))
# R-script to import immune Pathways which shall be highlighted in the publication
source(file.path("..", "..", "..", "r_scripts", "pathway_analysis", 'pathway_lists.R'))
# R-script to plot the Pathway enrichment result
source(file.path("..", "..", "..", "r_scripts", "pathway_analysis", "plot_pathways.R"))
```

Define input directory.
```{r}
## Input 
# Input directory 
# TODO change to relative path
# input.folder <- file.path("..", "..", "input", "pathway_analysis")
input.folder <- file.path("/Users/christina.hillig/PycharmProjects/ST_Immune_publication/Publication_analysis/output/reviewers/Figure_S8/glmGamPoi_DEGs")
# Date
date <- "2022-09-16"

cytokine <- 'IL17A'

input.dir <- file.path(input.folder, date, cytokine)
```
Determine cut-parameters to identify differential expressed genes and enriched pathways.
```{r}
# Cut parameter
l2fc.factor <- 1
fdr.value <- 0.05
pval.cut <- 0.05
p.value <- 0.05
minGSSize <- 10
# Multi-test method Benjamini-Hochberg (BH)
multitest.method <-  "BH"
```
```{r}
# Plot Parameters
show_dotplot_categories <- 15

width_img <- 16
height_img <- 8
```
Create output directory.
```{r}
## Output
# General output directory
output.dir <- file.path("/Users/christina.hillig/PycharmProjects/ST_Immune_publication/Publication_analysis/output/reviewers/Figure_S8/REACTOME_pathways", Sys.Date(), cytokine)
dir.create(output.dir, showWarnings = FALSE, recursive = TRUE)

# print(output.dir)
```
Load DGE analysis results and prepare dataframe for further analysis steps.
```{r}
# 1. Get all DGE .csv files in subfolders
all_filenames <- list.files(path = input.dir, pattern = ("*.csv"), recursive = TRUE)
# 2.1 remove metaData from list
dge_filename <- all_filenames[!grepl("metaData*", all_filenames)]
# 2.2 read out cytokine .csv file
dge_filename <- all_filenames[grepl(cytokine, all_filenames)]
# 3. Load DGE Analysis result file with colnames: 
# "X" "gene_symbol" "gene_name" "entrezid" "pval" "padj" "log2fc" "hkg"
df.dge_res <- load_files(path_name_file = file.path(input.dir,  dge_filename))
df.dge_res$X <- NULL

print.data.frame(df.dge_res[1:10, ], digits = 4)
```

# Pathway enrichment analysis
The pathway enrichment analysis included the following steps:
1. Identify significant and background genes 
1. Run PA analysis
1. Plot Save PA analysis results  

In a first step, we convert the gene symbol to entrezID. That is needed to use the later the ReactomePA::enrichPathway function. In best case use always the esemblID.
```{r}
df.dge_res <- rename_genetoentrezid(df.dge_results = df.dge_res)
```


## Get significantly DEx genes and background genes
Now we can start with idetifying differential expressed genes with our manually set cut parameters.
```{r}
# I. Define significant DEx genes and background genes
# I.a) sort genes into groups belonging either to reference or test (control) condition
df.ref_degenes <- get_significantgenes(df.dge_results = df.dge_res, p_value = p.value,
                              lfc_factor = l2fc.factor, op = `>`) 
df.ref <- df.ref_degenes[[1]]
# I. Rank genes based on their fold change
ranked_genes.ref <- do_rank_genes(df.dge_results = df.ref)

df.ctrl_degenes <- get_significantgenes(df.dge_results = df.dge_res, p_value = p.value,
                                       lfc_factor = -l2fc.factor, op = `<`) 
df.ctrl <- df.ctrl_degenes[[1]]

# I. Rank  genes based on their fold change
ranked_genes.ctrl <- do_rank_genes(df.dge_results = df.ctrl)

# I.b) Background genes are all genes from our (sup-) data set
bg_genes <- as.character(df.dge_res$entrezid)
```


## Identify enriched pathways using ReactomePA
```{r}
# II.a) Find enriched Pathways for reference condition
reactome_object.ref <- ReactomePA::enrichPathway(
  gene = df.ref_degenes[[2]], # a vector of entrezID
  universe = bg_genes, organism = 'human', 
  qvalueCutoff = fdr.value, pvalueCutoff = pval.cut,  pAdjustMethod = multitest.method, 
  minGSSize = minGSSize, maxGSSize = 500, readable = T)
```

In order to plot gene names instead of entrezIDs they have to be converted. For this use the function "*DOSE::setReadable*" to convert entrezIDs to gene symbol.
```{r}
################### ---> convert gene ID to Symbol <--- ################### 
reactome.ref <- setreadable_pa(paenrich_object = reactome_object.ref)
print("Pathways associated with cytokine-positive group")
print.data.frame(reactome.ref[1:3, ], digits = 4)
```


## Save and Plot Pathwyas
```{r}
#############################################################################
################### ---> Save results to csv file <--- ###################### 
#############################################################################
# Attention: 
# ctrl (= negative log2FC) and ref (= positive log2FC) are switched for Immune publication
save_enrichobject_as_csv(paenrich_object = reactome.ref, condition = cytokine, 
                         pa_database = 'REACTOME', output_path = output.dir) 
```

Visualise Pathways in a cnet- and dotplot.
First, we plot the enriched pathways of the *IL17A*-positive group and save the plots as .pdf.
```{r,  fig.width=width_img, dpi = 300 , fig.height=height_img}
# III.b) Reference Condition
# If a gene is associated with two or more enriched PAs 
# but less than those are shown than this results in a bug 
# ==> the log2FC of that gene is not correctly shown
if (!is.null(nrow(reactome.ref)))
{
  if (nrow(reactome.ref) > 1)
  {
    # Cnetplots to visualise enriched pathways
    fig.pathways.REACTOME(reactome_res = reactome.ref,
                          entrezid_log2fc = ranked_genes.ref,
                          showCategories = 3, 
                          sig.cytokine = cytokine,
                          output.dir = output.dir,
                          title = paste(
                            cytokine, "_REACTOME_Pathway_Enrichment_Analysis.pdf", sep=''), 
                          width = 12, height = 6)

    # Dotplot to visualise enriched pathways
    fig.pathway.dotplot(pathway_res = reactome.ref,
                        showCategories = 10,
                        method = 'REACTOME', 
                        output.dir = output.dir,
                        title = paste(cytokine, "_REACTOME_dotplot.pdf", sep=''), 
                        width = 6, height = 6)
  }
}

```


# zscore=(up−down) / sqrt(count)

# References
Cite used Bioconductor packages: 
```{r}
citation("ReactomePA")
citation("DOSE")
citation("org.Hs.eg.db")
citation("enrichplot")
citation("pathview")
```

```{r}
sessionInfo()
```



